Antigenic phenotypes and phylogenetic metrics improve forecasts of seasonal influenza A/H3N2 evolution

### 2019-10-22 manuscript/abstract.tex
Address two previously open questions
We decided to report both validation and test results and to report the
difference between models and the naive model as the primary statistic.
Include edits and comments

\begin{itemize}
\item{How should forecast uncertainty be calculated?}
\item{Should we calculate pairwise distances using a more evolutionarily-informed metric than just Hamming distance?}
\item{Why do all of the substitution-based fitness metrics perform so poorly (especially the epitope mutations) when epitope mutations have been previously associated with viral fitness?}
\item{How does our model performance compare with the original clade-based models?}
\item{Should we report both validation and test results? Should we report them together or separate?}
\item{Should we report difference from naive as a primary statistic? As naive - model or as model - naive?}
\end{itemize}

>>>>

\begin{itemize}
\item{How should forecast uncertainty be calculated? \tbc{Important and will be commented on by reviewers. I think this can be carried by purely retrospective comparisons however.}}
\item{Should we calculate pairwise distances using a more evolutionarily-informed metric than just Hamming distance? \tbc{I'm quite happy with Hamming distance.}}
\item{Why do all of the substitution-based fitness metrics perform so poorly (especially the epitope mutations) when epitope mutations have been previously associated with viral fitness? \tbc{This is still concerning. I wonder about doing a quick calculation of a tree that we output to count epitope mutations and non-epitope mutations to trunk vs side-branches. We should be able to replicate the previous results. Katherine was able to do this just now with modern data with the Wolf epitope sites.}}
\item{How does our model performance compare with the original clade-based models? \tbc{I think the clade performance figure is enough here.}}
\item{\tbc{Do we need to calculate LBI via a region-weighted scheme? May be better left for a later update? But seems important given centrality of LBI to our predictions.}}
\end{itemize}

### 2019-10-22 manuscript/main.tex
Include edits and comments

Seasonal influenza virus infects 5--15\% of the global population every year causing an estimated 250,000 to 500,000 deaths annually \cite{flufactsheet}.
Vaccination remains the most effective public health response available.
However, frequent viral mutation results in viruses that escape previously acquired human immunity.
The World Health Organization (WHO) selects vaccine viruses to match circulating viruses, but because the process of vaccine development and distribution requires several months to complete, accurate vaccine strain selection requires a prediction of which viruses will predominate approximately one year after vaccine viruses are selected.
Current vaccine predictions favor viruses that are distinct from prior vaccine viruses in the hemagglutinin (HA) protein, which acts as the primary target of human immunity.
The hemagglutination inhibition (HI) assay \cite{hirst1943studies} is used to measure the degree of cross-reactivity between pairs of circulating viruses.
HI assays are fundamental for vaccine strain selection, but they are laborious and low-throughput compared to genome sequencing \cite{Wood:2012ii}.
As a result, researchers have developed computational methods to predict influenza fitness from sequence data alone \cite{Luksza:2014hj,Steinbruck:2014kq,Neher:2014eu}.

@@

\begin{figure*}[ht]
  \begin{center}
  \includegraphics[width=\columnwidth]{figures/distance-based-fitness-model.png}
  \caption{
    Schematic representation of the fitness model for simulated A/H3N2-like populations wherein the fitness of samples at timepoint $t$ determines the estimated frequency of samples with similar sequences one year in the future at timepoint $u$.
    Samples are colored by their amino acid sequence composition such that genetically similar samples have similar colors.
    A) Samples at timepoint $t$ are shown in their phylogenetic context and sized by their frequency at that timepoint.
    The estimated future population at timepoint $u$ is projected to the right with samples scaled in size by their projected frequency based on the true fitness model.
    B) The frequency trajectories of samples at timepoint $t$ to $u$ represent the predicted the growth of the orange samples to the detriment of the purple samples.
    C) Samples at timepoint $u$ are shown in the corresponding phylogeny for that timepoint and scaled by their frequency at that time.
    D) The observed frequency trajectories of samples at timepoint $u$ broadly recapitulate the model's forecasts while also revealing increased diversity of sequences at the future timepoint that the model could not anticipate (e.g., the emergence of the yellow cluster from within the successful orange cluster).
  }
  \label{fig:model}
  \end{center}
\end{figure*}

Here, we present a model of seasonal influenza evolution inspired by the exponential growth model of {\L}uksza and L\"assig \cite{Luksza:2014hj}.
As with this original model, we seek to forecast the frequencies of viral populations one year in advance by applying to each virus sample an exponential growth factor scaled by an estimate of the sample's fitness (Fig.~\ref{fig:model}).
We estimate the frequency of virus samples every six months using a smoothed KDE kernel to represent the frequency of each sample.
We estimate viral fitness with biologically-informed metrics including those originally defined by \cite{Luksza:2014hj} of epitope cross-immunity and non-epitope mutations as well as four more recent metrics including hemagglutination inhibition (HI) cross-immunity \cite{Neher:2016hy}, deep mutational scanning (DMS) mutational effects \cite{Lee2018}, local branching index (LBI) \cite{Neher:2014eu}, and change in clade frequency over time (delta frequency).
We fit models by learning coefficients for each fitness metric either individually or in linear combinations from training data and select the best of these models using time-series cross-validation.
After selecting optimal models from training and validation, we evaluate the true out-of-sample errors of these models on additional data that were held out from the initial model fitting and tuning.
Importantly, our models find fitness coefficients that minimize the normalized average Hamming distance between the observed population one year in the future and the estimated population produced by the exponential growth model (Fig.~\ref{fig:model}).
With this approach, we avoid the intrinsic instability of clade definitions due to variability in phylogenetic reconstruction from year to year.
However, we retain the benefits of fitting models to highly similar samples found within clades and enable future forecasting efforts for pathogens whose sequences are not amenable to standard phylogenetic inference.

@@

\begin{figure*}[ht!]
  \begin{center}
  \includegraphics[width=\textwidth]{figures/unadjusted-model-accuracy-and-coefficients-for-simulated-populations.png}
  \caption{
    Model a) coefficients and b) distances closer to the future measured in amino acids (AAs) for each model relative to a naive model for simulated populations of A/H3N2-like viruses.
    Coefficients are shown for each individual fitness metric per validation timepoint (N=33) in solid circles with the mean $\pm$ standard deviation in the top-left corner of each panel.
    Distances closer to the future in amino acids are shown per validation or test timepoint for each individual model relative to the observed distance between timepoints.
    Models outperform the naive model when the distances closer to the future are greater than zero.
    The mean $\pm$ standard deviation of amino acids per validation timepoint are shown in the top-left of each panel.
    The same values are shown per test timepoint in the top-right of each panel.
    Solid circles indicate validation timepoints and open circles indicate test timepoints when coefficients are fixed to their mean values from training/validation and applied to out-of-sample test data.
  }
  \label{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}
  \end{center}
\end{figure*}

\begin{table*}[ht]
  \begin{center}
    \input{tables/simulated_model_selection.tex}
    \caption{
      Model performance with A/H3N2-like populations relative to the naive model.
      Models use one or more fitness metrics to minimize the distance between the population of HA amino acid sequences at timepoint, $t$, and those at a timepoint one year in the future, $t + 1$.
      The naive model assumes the populations at time $t$ and $t + 1$ are identical, effectively measuring the observed distance between the two timepoints.
      Better models produce estimates that are closer to the future population than the naive model.
    }
    \label{table_simulated_model_selection}
  \end{center}
\end{table*}

@@

\begin{figure*}[ht]
  \begin{center}
  \includegraphics[width=\textwidth]{figures/test-of-best-model-for-simulated-populations.png}
  \caption{
  Test of best model for simulated populations of A/H3N2-like viruses using 10 years previously unobserved test data and fixed model coefficients.
  a) The correlation of estimated and observed clade growth rates shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
  b) The rank of the estimated best strain based on its distance to the future in the best model shows how often the model makes a good choice when forced to select a single representative strain for the future population.
  }
  \label{fig:test_of_best_model_for_simulated_populations}
  \end{center}
\end{figure*}

@@

After identifying the best model to forecast natural populations, we integrated our forecasting model into our existing open source pathogen surveillance framework, Nextstrain \cite{Hadfield2018}, and used our model to forecast A/H3N2 populations in September 2020.
Our combined non-epitope mutations and LBI model favored the growth of the currently circulating clade A1b/197R, the maintenance of A1b/137F, and the decline of 3c3.A (Fig.~\ref{fig:nextstrain_forecasts}).

>>>>

Seasonal influenza virus infects 5--15\% of the global population every year causing an estimated 250,000 to 500,000 deaths annually \cite{flufactsheet}.
Vaccination remains the most effective public health response available.
However, frequent viral mutation results in viruses that escape previously acquired human immunity.
The World Health Organization (WHO) selects vaccine viruses to match circulating viruses, but because the process of vaccine development and distribution requires several months to complete, accurate vaccine strain selection requires a prediction of which viruses will predominate approximately one year after vaccine viruses are selected.
Current vaccine predictions focus on the hemagglutinin (HA) protein, which acts as the primary target of human immunity.
The hemagglutination inhibition (HI) assay \cite{hirst1943studies} is used to measure the degree of cross-reactivity between pairs of circulating viruses.
HI assays are fundamental for vaccine strain selection, but they are laborious and low-throughput compared to genome sequencing \cite{Wood:2012ii}.
As a result, researchers have developed computational methods to predict influenza fitness from sequence data alone \cite{Luksza:2014hj,Steinbruck:2014kq,Neher:2014eu}.

@@

\begin{figure*}[ht]
  \begin{center}
  \includegraphics[width=\columnwidth]{figures/distance-based-fitness-model.png}
  \caption{
    Schematic representation of the fitness model for simulated A/H3N2-like populations wherein the fitness of samples at timepoint $t$ determines the estimated frequency of samples with similar sequences one year in the future at timepoint $u$.
    Samples are colored by their amino acid sequence composition such that genetically similar samples have similar colors.
    A) Samples at timepoint $t$ are shown in their phylogenetic context and sized by their frequency at that timepoint.
    The estimated future population at timepoint $u$ is projected to the right with samples scaled in size by their projected frequency based on the true fitness model.
    B) The frequency trajectories of samples at timepoint $t$ to $u$ represent the predicted the growth of the orange samples to the detriment of the purple samples.
    C) Samples at timepoint $u$ are shown in the corresponding phylogeny for that timepoint and scaled by their frequency at that time.
    D) The observed frequency trajectories of samples at timepoint $u$ broadly recapitulate the model's forecasts while also revealing increased diversity of sequences at the future timepoint that the model could not anticipate, e.g. the emergence of the yellow cluster from within the successful orange cluster.
    \tbc{I'd label left and right columns as `Forecast' and `Retrospective' of something to this effect.}
  }
  \label{fig:model}
  \end{center}
\end{figure*}

Here, we present a model of seasonal influenza evolution inspired by the Malthusian growth fitness model of {\L}uksza and L\"assig \cite{Luksza:2014hj}.
As with this original model, we seek to forecast the frequencies of viral populations one year in advance by applying to each virus sample an exponential growth factor scaled by an estimate of the sample's fitness (Fig.~\ref{fig:model}).
We estimate the frequency of virus samples every six months using a smoothed KDE kernel to represent the frequency of each sample.
We estimate viral fitness with biologically-informed metrics including those originally defined by \cite{Luksza:2014hj} of epitope cross-immunity and non-epitope mutations as well as four more recent metrics including hemagglutination inhibition (HI) cross-immunity \cite{Neher:2016hy}, deep mutational scanning (DMS) mutational effects \cite{Lee2018}, local branching index (LBI) \cite{Neher:2014eu}, and change in clade frequency over time (delta frequency).
We fit models by learning coefficients for each fitness metric either individually or in linear combinations from training data and select the best of these models using time-series cross-validation.
After selecting optimal models from training and validation, we evaluate the true out-of-sample errors of these models on additional data that were held out from the initial model fitting and tuning.
Importantly, our models find fitness coefficients that minimize the normalized average Hamming distance between the observed population one year in the future and the estimated population produced by the exponential growth model (Fig.~\ref{fig:model}).
\tbc{I'd generally use the term `fitness model' instead of `exponential growth' model. Exponential growth makes me think of actual exponential growth rather than Malthusian growth.}
With this approach, we avoid the intrinsic instability of clade definitions due to variability in phylogenetic reconstruction from year to year.
However, we retain the benefits of fitting models to highly similar samples found within clades and enable future forecasting efforts for pathogens whose sequences are not amenable to standard phylogenetic inference.

@@

\begin{figure*}[ht!]
  \begin{center}
  \includegraphics[width=\textwidth]{figures/unadjusted-model-accuracy-and-coefficients-for-simulated-populations.png}
  \caption{
    Model a) coefficients and b) distances closer to the future measured in amino acids (AAs) for each model relative to a naive model for simulated populations.
    Coefficients are shown for each individual fitness metric per validation timepoint (N=33) in solid circles with the mean $\pm$ standard deviation in the top-left corner of each panel.
    Distances closer to the future in amino acids are shown per validation or test timepoint for each individual model relative to the observed distance between timepoints.
    Models outperform the naive model when the distances closer to the future are greater than zero.
    The mean $\pm$ standard deviation of amino acids per validation timepoint are shown in the top-left of each panel.
    The same values are shown per test timepoint in the top-right of each panel.
    Solid circles indicate validation timepoints and open circles indicate test timepoints when coefficients are fixed to their mean values from training/validation and applied to out-of-sample test data.
  }
  \label{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}
  \end{center}
\end{figure*}

\begin{table*}[ht]
  \begin{center}
    \input{tables/simulated_model_selection.tex}
    \caption{
      Model performance with simulated populations relative to the naive model.
      Models use one or more fitness metrics to minimize the distance between the population of HA amino acid sequences at timepoint, $t$, and those at a timepoint one year in the future, $t + 1$.
      The naive model assumes the populations at time $t$ and $t + 1$ are identical, effectively measuring the observed distance between the two timepoints.
      Better models produce estimates that are closer to the future population than the naive model.
      \tbc{Confused as to why `epitope cross-immunity' isn't working in the simulated example.}
    }
    \label{table_simulated_model_selection}
  \end{center}
\end{table*}

@@

\begin{figure*}[ht]
  \begin{center}
  \includegraphics[width=\textwidth]{figures/test-of-best-model-for-simulated-populations.png}
  \caption{
  Test of best model for simulated populations using 10 years previously unobserved test data and fixed model coefficients.
  a) The correlation of estimated and observed clade growth rates shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
  b) The rank of the estimated best strain based on its distance to the future in the best model shows how often the model makes a good choice when forced to select a single representative strain for the future population.
  }
  \label{fig:test_of_best_model_for_simulated_populations}
  \end{center}
\end{figure*}

@@

After identifying the best model to forecast natural populations, we integrated our forecasting model into our existing open source pathogen surveillance framework, Nextstrain \cite{Hadfield2018}, and used our model to forecast A/H3N2 populations in September 2020.
Our combined non-epitope mutations and LBI model favored the growth of the currently circulating clade A1b/197R, the maintenance of A1b/137F, and the decline of 3c3.A (Fig.~\ref{fig:nextstrain_forecasts}).
\tbc{Include figure showing predicted match of strains to future population. Possibly compare Hamming match to FRA match.}
